The evolution of gene expression in the term placenta of viviparous mammals

### 2015-08-28 manuscript/manuscript.Rnw
add more missing references to the manuscript

@@

@@

The top 50 expressed genes in each genome analyzed are shown in Tables
2-7, and all genes for each transcriptome are listed in Supplementary
Tables X-XX [These are large Excel files to be included with the
Supplemental Material]. In addition, specific genes including
placental hormones and hormone-related genes, galectins, and
insulin-like growth factor related genes are shown with FPKM values
side to side for all transcriptomes including the reanalysis of mouse
from Wang et al. (2011) in Table 8. Below we describe each
transcriptome separately, specifically concentrating on
highly-expressed genes.

@@

Reproductive hormones are central to the regulation of pregnancy and
their source, timing, and quantity can vary widely from species to
species (Carter, 2012). Expression and synthesis of hormones is
perhaps the most well-documented physiological process in the placenta
in many species. Here we quantified the extreme disparity in genes
with hormone products or enzymatic products that directly produce
hormones in multiple mammalian species (Table 8).

In both ruminants and primates, the placenta is responsible for the
production of estrogen and progesterone. High relative expression of
CYP19A1, the enzyme that synthesizes estrogen, and HSD3B1, the enzyme
that synthesizes progesterone, in the human, bonobo, and cow,
generally matches the high levels of placental production of estrogen
and progesterone in humans, other primates, and ruminants in late
pregnancy. In addition, HSD3B1 shows high levels in dog, which matches
the high levels of placental production of progesterone at term in
that species.

The elephant is strikingly the only mammal analyzed that shows high
expression of the primarily pituitary hormone prolactin (PRL) in the
placenta. All afrotherians yet analyzed contain only one prolactin
gene, and prolactin has undergone adaptive evolution on the elephant
lineage (Wallis, 2009). (Yamamoto et al., 2011) noted the presence of
a prolactin in the trophoblast of the elephant placenta using
immunohistological staining and human derived antibodies. A previous
analysis of the elephant placenta also detected the expression of
prolactin (Hou et al., 2012). Here we confirm that the sole prolactin
in the elephant is not only expressed but highly expressed in the
placenta, and likely functions both in the maintenance of the corpus
luteum and as a placental lactogen (Yamamoto et al., 2011). PRL is not
expressed in the opossum, even though there is evidence that it is
expressed in the placenta of another marsupial, the tammar wallaby
(Menzies et al., 2011). This suggests that placental prolactin
expression in marsupials may be as diverse as in eutherian mammals.
Although, PRL is not expressed to any great degree in the placenta
other mammals, we see very high expression of genes which resulted
from duplications of PRL in both the cow (Table X), as well as mouse
(Supplemental Table X). Describe these

The growth hormone (GH) family has expanded in anthropoid primates,
resulting in approximately 5 genes in humans (GH1, GH2, CSH1, CSH2,
CSHL1). All but GH1 show substantial expression in the placenta (Table
8). There is some evidence that increased expression of these genes
may be correlated with overall birth weight of offspring in humans
(Mannik et al., 2010), and it is possible that the extreme expression
levels of these genes may be due to high birth weight. Ensembl
annotations show 4 GH genes present in the chimpanzee genome, and
three of these genes show high expression in bonobo, although to a
lesser degree than in human (Table 8). Human placental GH genes are
involved in increasing blood flow, replacing maternal pituitary
expression of GH, as well as trophoblast invasion (Alsat et al., 1998;
Lacroix et al., 2005; Schiessl et al., 2007), and it is likely that
they have the same function in the bonobo.

The human gene CGA codes for the alpha subunit of multiple hormones
including pituitary hormones (luteinizing hormone,
follicle-stimulating hormone, thyroid-stimulating hormone), as well as
numerous chorionic gonadotropins with variable beta subunits (Maston
and Ruvolo, 2002). The function of the chorionic gonadotropins is to
prevent the regression of the corpus luteum (Hanson et al., 1971).
Among the six placentas we analyzed here, CGA is only expressed in the
placentas of humans and bonobos (Table 8), which corresponds with the
emergence of the chorionic gonadotropins in anthropoid primates
(Maston and Ruvolo, 2002). However, oddly, CGA is expressed at much
higher levels in both species than the beta subunits with which its
product dimerizes (Table 8). CGA is one of several genes noted by
(Uddin et al., 2008) to have undergone adaptive evolution of the
human-chimpanzee stem lineage.

Relaxin is a hormone that has multiple effects during pregnancy
including increasing maternal blood flow, widening the pubic bone and
relaxing uterine musculature in preparation for labor. In both dogs
and elephants, there is high expression of relaxin in the term
placenta, with little evidence in other species examined here (Table
8). Relaxin is known to be produced in the placenta of the pregnant
dog and is used in the diagnosis of canine pregnancy (Steinetz et al.,
1989; Tsutsui and Stewart, 1991); however, the source of relaxin
produced by pregnant African elephants remains unclear (Meyer et al.,
2004). The high rise in relaxin in elephants has been suggested to be
a major signal for the initiation of labor (Niemuller et al., 1998).
Here we show that the placenta of elephants show high expression of
relaxin and is likely the predominant source of relaxin at term (Table
8).

>>>>

\DLA{Complete discussion of the ateles placenta}

@@

\DLA{Need to do an embargoed submission to SRA}

@@

\DLA{Summarize all of this in a table}

The top 50 expressed genes in each genome analyzed are shown in
\DLA{Tables 2-7}, and all genes for each transcriptome are listed in
\DLA{Supplementary Tables X-XX} [These are large Excel files to be
included with the Supplemental Material]. In addition, specific genes
including placental hormones and hormone-related genes, galectins, and
insulin-like growth factor related genes are shown with FPKM values
side to side for all transcriptomes including the reanalysis of mouse
from \cite{Wang.ea2011:surveyfornovelimprinted} in \DLA{Table 8}.
Below we describe each transcriptome separately, specifically
concentrating on highly-expressed genes.

@@

Reproductive hormones are central to the regulation of pregnancy and
their source, timing, and quantity can vary widely from species to
species \cite{Carter2012:Evolutionofplacentalfunction}. Expression and synthesis of hormones is
perhaps the most well-documented physiological process in the placenta
in many species. Here we quantified the extreme disparity in genes
with hormone products or enzymatic products that directly produce
hormones in multiple mammalian species (\DLA{Table 8}).

In both ruminants and primates, the placenta is responsible for the
production of estrogen and progesterone. High relative expression of
CYP19A1, the enzyme that synthesizes estrogen, and HSD3B1, the enzyme
that synthesizes progesterone, in the human, bonobo, and cow,
generally matches the high levels of placental production of estrogen
and progesterone in humans, other primates, and ruminants in late
pregnancy. In addition, HSD3B1 shows high levels in dog, which matches
the high levels of placental production of progesterone at term in
that species. \DLA{I think we need citations for this}

The elephant is strikingly the only mammal analyzed that shows high
expression of the primarily pituitary hormone prolactin (PRL) in the
placenta. All afrotherians yet analyzed contain only one prolactin
gene, and prolactin has undergone adaptive evolution on the elephant
lineage \cite{Wallis2009:ProlactininAfrotheriacharacterization}.
\cite{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
presence of a prolactin in the trophoblast of the elephant placenta
using immunohistological staining and human derived antibodies. A
previous analysis of the elephant placenta also detected the
expression of prolactin (Hou et al., 2012). Here we confirm that the
sole prolactin in the elephant is not only expressed but highly
expressed in the placenta, and likely functions both in the
maintenance of the corpus luteum and as a placental lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}. PRL is not expressed in the opossum, even though there
is evidence that it is expressed in the placenta of another marsupial,
the tammar wallaby \cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests that
placental prolactin expression in marsupials may be as diverse as in
eutherian mammals. Although, PRL is not expressed to any great degree
in the placenta other mammals, we see very high expression of genes
which resulted from duplications of PRL in both the cow \DLA{(Table X)}, as
well as mouse \DLA{(Supplemental Table X)}. \DLA{Describe these}

The growth hormone (GH) family has expanded in anthropoid primates,
resulting in approximately 5 genes in humans (GH1, GH2, CSH1, CSH2,
CSHL1). All but GH1 show substantial expression in the placenta (Table
8). There is some evidence that increased expression of these genes
may be correlated with overall birth weight of offspring in humans
\DLA{(Mannik et al., 2010)}, and it is possible that the extreme expression
levels of these genes may be due to high birth weight. Ensembl
annotations show 4 GH genes present in the chimpanzee genome, and
three of these genes show high expression in bonobo, although to a
lesser degree than in human (Table 8). Human placental GH genes are
involved in increasing blood flow, replacing maternal pituitary
expression of GH, as well as trophoblast invasion \cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth}, and it is likely that
they have the same function in the bonobo.

The human gene CGA codes for the alpha subunit of multiple hormones
including pituitary hormones (luteinizing hormone,
follicle-stimulating hormone, thyroid-stimulating hormone), as well as
numerous chorionic gonadotropins with variable beta subunits \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of the chorionic gonadotropins is to
prevent the regression of the corpus luteum \cite{Hanson.ea1971:EffectsofHCGand}.
Among the six placentas we analyzed here, CGA is only expressed in the
placentas of humans and bonobos (\DLA{Table 8}), which corresponds with the
emergence of the chorionic gonadotropins in anthropoid primates
\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly, CGA is expressed at much
higher levels in both species than the beta subunits with which its
product dimerizes (\DLA{Table 8}). CGA is one of several genes noted by
\cite{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone adaptive evolution of the
human-chimpanzee stem lineage.

Relaxin is a hormone that has multiple effects during pregnancy
including increasing maternal blood flow, widening the pubic bone and
relaxing uterine musculature in preparation for labor. In both dogs
and elephants, there is high expression of relaxin in the term
placenta, with little evidence in other species examined here (Table
8). Relaxin is known to be produced in the placenta of the pregnant
dog and is used in the diagnosis of canine pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof};
however, the source of relaxin produced by pregnant African elephants
remains unclear \cite{Meyer.ea2004:Speciesandfetalgender}. The high
rise in relaxin in elephants has been suggested to be a major signal
for the initiation of labor
\cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive} Here we
show that the placenta of elephants show high expression of relaxin
and is likely the predominant source of relaxin at term \DLA{(Table 8)}.
